---
title: "Understanding World Population Dynamics"
subtitle: "Assignment 1 - PSYC593"
format:
   html: 
    output-file: "report"
    output-ext:  "html"
project:
  output-dir: '/output'
date: "09-05-2023"
author: "Jacqueline Erens"
---

<!-- These problems come from Imai, K. & Webb Williams, N. (2018). Quantitative Social Science: An introduction -->
<!-- in tidyverse. Princeton University Press. -->

<!-- The code giving partial solutions by Víctor H Cervantes -->

Understanding population dynamics is important for many areas of
social science.  We will calculate some basic demographic quantities
of births and deaths for the world's population from two time periods:
1950 to 1955 and 2005 to 2010.  We will analyze the following CSV data
files - `Kenya.csv`, `Sweden.csv`, and `World.csv`.  Each file
contains population data for Kenya, Sweden, and the world,
respectively.  The table below presents the names and descriptions of
the variables in each data set.

<!-- Note the syntax to create a table in markdown -->
<!-- Before the table there must be an empty line, and -->
<!-- the table begins with a line of dashes. -->
<!-- The column names are separated with spaces --> 
<!-- and the first character must be a space -->
<!-- Then, segments of repeated dashes indicate the column separations (note -->
<!-- the space between groups of dashes before the character D of the column Description) -->
#This all looks good and according to format. Space is present in the place where there'd be a line if this was like an excel table
#Do people care about how long the dashes go for, or is it just in line with the longest column of text

--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `country`            Abbreviated country name
 
 `period`             Period during which data are collected
 
 `age`                Age group 
 
 `births`             Number of births in thousands (i.e., number of children 
                      born to women of the age group) 
 
 `deaths`             Number of deaths in thousands 
 
 `py.men`             Person-years for men in thousands 
 
 `py.women`           Person-years for women in thousands
--------------------------------------------------------------------------------

Source: United Nations, Department of Economic and Social Affairs, 
Population Division (2013). *World Population Prospects: The 2012 Revision, DVD Edition.*

```{r load packages, echo = FALSE}
# Load packages ----
library(tidyverse)
```

```{r read in data, echo = FALSE}
# Read data ----
# Paths were changed to the appropriate path for the current working directory structure, without specifying absolute paths, as to avoid errors
world_data  <- readr::read_csv("../data/World.csv")
kenya_data  <- readr::read_csv("../data/Kenya.csv")

# This should be changed from SwedenData to sweden_data to be consistent with Snake Case (following first two)
sweden_data <- readr::read_csv("../data/Sweden.csv")
```

<!-- The `read_csv` function from the package `readr` -->
<!-- is part of the `tidyverse`. It will read the data into a -->
<!-- special `data.frame` with more structure known as a `tibble` -->

The data are collected for a period of 5 years where *person-year* is
a measure of the time contribution of each person during the
period. For example, a person that lives through the entire 5 year
period contributes 5 person-years whereas someone who only lives
through the first half of the period contributes 2.5 person-years.
Before you begin this exercise, it would be a good idea to directly
inspect each data set.  In R, this can be done with the `View`
function, which takes as its argument the name of a `data.frame` to be
examined.  Alternatively, in RStudio, double-clicking a `data.frame`
in the `Environment` tab will enable you to view the data in a
spreadsheet-like view.

```{r view all data}
# Instructions suggest visually inspecting the data using the View() function
View(world_data)
View(kenya_data)
View(sweden_data)
```

There is at the very least, no missing data. 
There were some years and age ranges where the number of births was 0.00, which I found odd at first and then remebered how biology works.
We shall proceed.

## Question 1

We begin by computing *crude birth rate* (CBR) for a given
  period.  The CBR is defined as:
  $$ 
    \text{CBR} 
     =  \frac{\text{number of births}}{\text{number of person-years lived}}
  $$
  <!-- It is also possible to include $\LaTeX$ syntax directly in the document. -->
  
  Compute the CBR for each period, separately for Kenya, Sweden, and
  the world.  Start by computing the total person-years, recorded as a
  new variable within each existing `data.frame` via the
  `$` operator, by summing the person-years for men and women.
  
  Then, store the results as a vector of length 2 (CBRs for two
  periods) for each region with appropriate labels.  You may wish to
  create your own function for the purpose of efficient programming.
  Briefly describe patterns you observe in the resulting CBRs.
  
### Answer 1

```{r calculate total person years}
# Create new variable py = total person years for each data set. Add spaces between operators. I chose to add extra for alignment, unclear if that helps or not
world_data$py  <-  world_data$py.men +  world_data$py.women
kenya_data$py  <-  kenya_data$py.men +  kenya_data$py.women
sweden_data$py <- sweden_data$py.men + sweden_data$py.women
```


<!-- Whenever you find that you will repeat the same calculation -->
<!-- on different sets of data, **consider** creating a function -->
<!-- instead of repeating the code for each piece of data. -->

Write reusable function to calculate CBR.

```{r CBR function}
# Function to compute the Crude Birth Rate (CBR). Spaces added to separate pipes, removed from between function and (). populationData converted to Snake Case for formatting.
compute_cbr <- function(population_data) {
  population_data %>% 
    group_by(period) %>% 
      summarise(cbr = sum(births) / sum(py)) %>% 
        pull()
}
```

<!-- The "pipe" `%>%` from package `magrittr` takes the output -->
<!-- from the left hand side and makes it the first argument of -->
<!-- the expression on the right hand side. -->

<!-- Starting in R 4.1.0, there is a base pipe operator `|>` -->
<!-- which has a similar behavior. -->

```{r CBR computation}
# Compute the CBR for each data set. Spaces were added after <- operator to conform with Tidyverse Style
(world_cbr  <- compute_cbr(world_data))
(kenya_cbr  <- compute_cbr(kenya_data))
(sweden_cbr <- compute_cbr(sweden_data))
```

With some minor changes to formatting of variable names (sweden_data) getting carried through, we get 3 vectors of 2 values.

The CBR is higher for kenya than Sweden at both time periods. 
The rate for Sweden is lower than the world rate, while Kenya's rate is higher, for both time periods
The rate dips a bit for everyone in the second period compared to the first.
  
## Question 2
  
The CBR is easy to understand but contains both men and women
   of all ages in the denominator. We next calculate the *total
     fertility rate* (TFR). Unlike the CBR, the TFR adjusts for age
   compositions in the female population.  To do this, we need to
   first calculate the *age specific fertility rate* (ASFR), which
   represents the fertility rate for women of the reproductive age
   range $[15, 50)$.  The ASFR for age range $[x, x+\delta)$, where
   $x$ is the starting age and $\delta$ is the width of the age range
   (measured in years), is defined as:
  $$
    \text{ASFR}_{[x,\ x+\delta)} 
    \ = \ \frac{\text{number of births to women of age $[x,\ x+\delta)$}}{\text{Number of person-years lived by women of age $[x,\ x+\delta)$}}
  $$
  Note that square brackets, $[$ and $]$, include the limit whereas
  parentheses, $($ and $)$, exclude it.  For example, $[20, 25)$
  represents the age range that is greater than or equal to 20 years
  old and less than 25 years old.  In typical demographic data, the
  age range $\delta$ is set to 5 years.  Compute the ASFR for Sweden
  and Kenya as well as the entire world for each of the two periods.
  Store the resulting ASFRs separately for each region.  What does the
  pattern of these ASFRs say about reproduction among women in Sweden
  and Kenya?
  

### Answer 2

Write a reusable function to compute ASFR.

```{r reusable asfr function}
# Function to compute Age specific fertility rate (ASFR), some spacing was modified to conform with guidelines
compute_asfr <- function(pop_data) {
  pop_data %>% 
    mutate(asfr = births / py.women)
}
```

Compute the values for the world & countries.

```{r compute asfr}
# Compute ASFR for each data set
world_data  <- compute_asfr(world_data)
kenya_data  <- compute_asfr(kenya_data)
sweden_data <- compute_asfr(sweden_data)
```

Display data for Kenya and Sweden to compare rates.

```{r display asfr}
# Compare ASFRs for Kenya and Sweden
kenya_data$asfr
sweden_data$asfr
```

This all seems okay, and it has gone into the overall data frame in a way that looks sensible.

```{r difference}
# Subtract the vector for Sweden's asfr from Kenya's asfr to check if Kenya's is truly always a greater number
diff_kenya_sweden_asfr <- kenya_data$asfr - sweden_data$asfr
diff_kenya_sweden_asfr
```

The ASFR is always higher for Kenya than Sweden, sometimes by more than double.
You can see this in the two vector output just via observation,
and can confirm by making a vector that is the difference of the two.
The numbers are always positive, indicating a magnitude difference (Kenya's asfr always is larger than Sweden's).

```{r difference as fraction}
# Make a ratio to see just how different, as decimals can sometimes be diffiuclt to interpret
frac_diff_kenya_sweden_asfr <- kenya_data$asfr / sweden_data$asfr
frac_diff_kenya_sweden_asfr
```

We can see that the difference is sometimes not that large, only about 2x the difference, 
and sometimes very large, approximately 62x the difference.

## Question 3

Using the ASFR, we can define the TFR as the average number of
  children women give birth to if they live through their entire
  reproductive age.
  $$
  \text{TFR} 
   =   \text{ASFR}_{[15,\ 20)} \times 5 + \text{ASFR}_{[20,\ 25)} \times 5 
  + \dots + \text{ASFR}_{[45,\ 50)} \times 5
  $$
  
  We multiply each age-specific fertility rate rate by 5 because the
  age range is 5 years.  Compute the TFR for Sweden and Kenya as well
  as the entire world for each of the two periods.  As in the previous
  question, continue to assume that women's reproductive age range is
  $[15, 50)$.  Store the resulting two TFRs for each country or the
  world as a vector of length two. 
  
  In general, how has the number of
  women changed in the world from 1950 to 2000? What about the total
  number of births in the world?


### Answer 3

Create the reusable function to calculate TFR.

```{r create function to compute TFR}
# Function to compute the total fertility rate (TFR)
compute_tfr <- function(population_data) {
  population_data %>% 
    group_by(period) %>% 
      summarise(tfr = 5 * sum(asfr) ) %>% 
        pull()
}
```

Apply it to the data frames (separately).

```{r use function to compute TFR}
# Compute the TFR for each data set
(world_tfr  <- compute_tfr(world_data))
(kenya_tfr  <- compute_tfr(kenya_data))
(sweden_tfr <- compute_tfr(sweden_data))
```

Divide the value in row 2 by the value in row 1 (total births by total women) to get this ratio.

```{r calculate changes}
# Compute totals of women and births in the world by period
(
world_data %>% 
  group_by(period) %>% 
  summarise(total_women  = sum(py.women),
            total_births = sum(births)) 
-> totals_world
)

# Check the output
totals_world 

# Compare how much these totals have changed. -1 used like this drops the first column in the dataframe, removing it from consideration.
(changes_totals <- totals_world[2, -1] / totals_world[1, -1])
```

I didn't realize you could use the -> operator at the end of a pipe like that, but that is neat!

The total number of women is increasing by a rate of about 2.5, and the total births is increasing by a rate of ~1.4

## Question 4

Next, we will examine another important demographic process:
  death.  Compute the *crude death rate* (CDR), which is a concept
  analogous to the CBR, for each period and separately for each
  region.  Store the resulting CDRs for each country and the world as
  a vector of length two. The CDR is defined as:
  $$ 
    \text{CDR} 
     =  \frac{\text{number of deaths}}{\text{number of person-years lived}}
  $$
  Briefly describe patterns you observe in the resulting CDRs.

Reusable function to calculate CDR.

```{r}
# Function to compute the Crude death rate (CDR)
compute_cdr <- function(population_data) {
  population_data %>% 
    group_by(period) %>% 
      summarise(cbr = sum(deaths) / sum(py)) %>% 
        pull()
}
```

Compute the CDRs for each data set and dispay.

```{r}
# Compute the CDR for each data set
(world_cdr  <- compute_cdr(world_data))
(kenya_cdr  <- compute_cdr(kenya_data))
(sweden_cdr <- compute_cdr(sweden_data))
```

The values for Kenya almost pattern exactly with the world ones.
Sweden is lower than the world values for the first periord, but very close to Kenya and the world values for the second period.


## Question 5

One puzzling finding from the previous question is that the CDR
  for Kenya during the period of 2005-2010 is about the same level as
  that for Sweden.  We would expect people in developed countries like
  Sweden to have a lower death rate than those in developing countries
  like Kenya.  While it is simple and easy to understand, the CDR does
  not take into account the age composition of a population.  We
  therefore compute the *age specific death rate* (ASDR).  The
  ASDR for age range $[x, x+\delta)$ is defined as:
  $$
    \text{ASDR}_{[x,\ x+\delta)} 
    \ = \ \frac{\text{number of deaths for people of age $[x,\ x+\delta)$}}
               {\text{number of person-years of people of age $[x,\ x+\delta)$}}
  $$
  Calculate the ASDR for each age group, separately for Kenya and
  Sweden, during the period of 2005-2010.  Briefly describe the
  pattern you observe.
 

Write reusable function to compute ASDR.

```{r}
# Function to compute Age specific death rate (ASDR). Indents fixed and population_data name used. Spaced out around operators
compute_asdr <- function(population_data) {
  population_data %>% 
    mutate(asdr = deaths / py)
}
```

Apply reusable function, add () around call to observe in output.

```{r}
# Compute ASDR for each data set. Sweden variable renamed.
(world_data  <- compute_asdr(world_data))
(kenya_data  <- compute_asdr(kenya_data))
(sweden_data <- compute_asdr(sweden_data))
```

I didn't expect those to be a data frame when the others were vectors. These values are all incredibly small and difficult to compare.

Maybe make ratios to compare them to each other?

```{r}
# Output vector of ratios making the comparisons to make observations
kenya_data$asdr  / world_data$asdr

sweden_data$asdr / world_data$asdr

kenya_data$asdr  / sweden_data$asdr
```
This wasn't as helpful as I thought, but I think the last one is what we want (comparing Kenya to Sweden).

Kenya's Age Specific Death Rate is often higher by either a slight bit or a great deal from Sweden's. But the value above 1 suggests it is always larger.

## Question 6

One way to understand the difference in the CDR between Kenya
  and Sweden is to compute the counterfactual CDR for Kenya using
  Sweden's population distribution (or vice versa).  This can be done
  by applying the following alternative formula for the CDR.
  $$
    \text{CDR}
    \ = \ \text{ASDR}_{[0, 5)} \times P_{[0,5)} + \text{ASDR}_{[5, 10)}
    \times P_{[5, 10)} + \cdots 
  $$
  where $P_{[x, x+\delta)}$ is the proportion of the population in the
  age range $[x, x+\delta)$.  We compute this as the ratio of
  person-years in that age range relative to the total person-years
  across all age ranges.  To conduct this counterfactual analysis, we
  use $\text{ASDR}_{[x,x+\delta)}$ from Kenya and $P_{[x,x+\delta)}$
  from Sweden during the period of 2005--2010.  That is, first
  calculate the age-specific population proportions for Sweden and
  then use them to compute the counterfactual CDR for Kenya. How does
  this counterfactual CDR compare with the original CDR of Kenya?
  Briefly interpret the result.



First, write a reusable function to compute the age specific population proportion.

```{r}
# Function to compute population proportion by period. I renamed popP because I liked pop_prop better. The last 3 lines will not nest indented, so I left it?
compute_pop_prop <- function(population_data) {
  population_data %>% 
    group_by(period) %>%
    mutate(pop_prop = py / sum(py)) %>%
    ungroup()
}
```

Next, apply that function to each dataset and assign it.

```{r}
# Compute population proportion for each data set. Spaces before assignment. Add () to display output
(world_data  <- compute_pop_prop(world_data))
(kenya_data  <- compute_pop_prop(kenya_data))
(sweden_data <- compute_pop_prop(sweden_data))
```

Now use the population proportions to calculate CDR.

```{r}
# Compute Kenyas CDR Kenya had Sweden's population distribution. Sweden data renamed, don't understand where "re" comes from in cdr_re_sweden. Relative to? Should we rename?
mutate(kenya_data,
 temp_cdr = asdr * sweden_data$pop_prop) %>%
  group_by(period) %>%
  summarise(cdr_re_sweden = sum(temp_cdr))
```

In comparisons to Sweden, the first period is about the same, the second period is double the rate.
